---
title: "Mandatory assignment 2"
author: "Jonas Øren"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(cache = TRUE)
knitr::opts_chunk$set(warnings = FALSE)
knitr::opts_chunk$set(messages = FALSE)
```

```{r packpages}
library(dplyr)
```



# Problem 1: Regression
We read the data
```{r}
daphnia_data <- read.csv("https://archive.ics.uci.edu/ml/machine-learning-databases/00505/qsar_aquatic_toxicity.csv", sep=';', header = F)
# Name the variables. The last column, LC50, is the response
names(daphnia_data) <- c("TPSA", "SAacc", "H050", "MLOGP", "RDCHI", "GATS1p",
                         "nN", "C040", "LC50")
n <- nrow(daphnia_data)
```

And split into a training and a test set

```{r train_test}
set.seed(1)
train_idx <- sample(1:n, size = round(2/3*n), replace = FALSE)
test_idx <- (1:n)[-train_idx]
```

## 1. Analysis of count variables

### Linear effect of count variables
```{r linear_effect}
# fit <- lm(LC50 ~ H050 + nN + C040, data = daphnia_data[train_idx, ])
fit_linear <- lm(LC50 ~ ., data = daphnia_data[train_idx, ])

```

### Dichotmoized count variables
```{r}
# Dochotomize count data to 0 (not present), and 1 (present)
dich_data <- daphnia_data
dich_data <- mutate_at(dich_data, c("H050", "nN", "C040"),
                       function(x) {x[x > 0] <- 1; return(factor(x))}
                       )
fit_dich <- lm(LC50 ~ ., data = dich_data[train_idx, ])
```

### Summaries
#### Coefficients
We first compare the two models by inspecting coefficient of the fitted regression models.

For the model with linear effects we have.


```{r}
summary(fit_linear)
```


The model with dichotomized count variables gives the following output.


```{r}
summary(fit_dich)
```
From the printouts we see that the linear effect of nN is significant while the dichotomized variable is not.

We also see that the coefficient for GATS1p in the dichotomized model have lower p-values than the coefficients in the
model with linear effect of all variables. This effect changes between different test and training set combinations, and
seem to be an effect of random variation more than from the model choice.


#### Prediction error as MSE in test data
We now inspect the prediction error, measured as squared error on the test data set
\[
 \frac{1}{n}\sum_{i \in \mathcal{T}}(\hat{y}_i - y_i)^2,
\]
where $\mathcal{T}$ is the set of indexes for the observations of the test set.



```{r}
# True respons on the test set
truth <- daphnia_data$LC50[test_idx]
# Predictions on the test set
prediction_linear_model <- predict(fit_linear, newdata = daphnia_data[test_idx, ])
prediction_dich_model <- predict(fit_dich, newdata = dich_data[test_idx, ])
cat(paste("Prediction error linear model:     ",
          round(mean((prediction_linear_model - truth)^2)
                               , 2), "\n" ) )
cat(paste("Prediction error dichotomous model:",
          round(mean((prediction_dich_model - truth)^2)
                               , 2) , "\n" ) )
```
We see that the model with linear effects for all the variables has a smaller prediction error on our test set.

From this analysis linear effects of count variables appear unambiguously better at predicting and explaining the outcome,
meaning the amount of molecules also has an effect, and we should not just look at the presence of molecules.


## 2. Repeating the procedure for different test and training sets
```{r}
set.seed(2)
# We will sum up the errors to compute the average
lin_mod_error_sum <- 0
dic_mod_error_sum <- 0

for (i in 1:200) {
  # Create new test and training sets
  train_idx_i <- sample(1:n, size = round(2/3*n), replace = FALSE)
  test_idx_i <- (1:n)[-train_idx_i]
  
  # Fit the new models
  fit_linear <- lm(LC50 ~ ., data = daphnia_data[train_idx_i, ])
  fit_dich <- lm(LC50 ~ ., data = dich_data[train_idx_i, ])
  
  # Find prediction error
  
  truth <- daphnia_data$LC50[test_idx_i]
  prediction_linear_model <- predict(fit_linear, newdata = daphnia_data[test_idx_i, ])
  prediction_dich_model <- predict(fit_dich, newdata = dich_data[test_idx_i, ])
  
  lin_mod_error_sum <- lin_mod_error_sum + mean((prediction_linear_model - truth)^2)
  dic_mod_error_sum <- dic_mod_error_sum + mean((prediction_dich_model - truth)^2)
}

cat(paste("Average prediction error linear model:     ",
          round(lin_mod_error_sum/200, 3), "\n",
          "Average prediction error dichotomous model:",
          round(dic_mod_error_sum/200, 3), "\n"))
```

The model with dichotomized effects has on average a bigger prediction error than the model with linear effects.
Dichotomizing the variables ignore the size of the effect, which discards much information. I would guess this will
be responsible for the poorer prediction performance of dichotomized variables.

## 3. Variable selection

http://www.sthda.com/english/articles/37-model-selection-essentials-in-r/154-stepwise-regression-essentials-in-r/

### Stepwise selection

```{r}
full_model <- lm(LC50 ~ ., data = daphnia_data[train_idx, ])
n <- length(train_idx)

forwards_AIC <- step(full_model, 
                     direction = "forward", 
                     k = 2, # AIC
                     trace = 0
                     )
backwards_AIC <- step(full_model, 
                     direction = "backward",
                     k = 2,
                     trace = 0
                     )
forwards_BIC <- step(full_model, 
                     direction = "forward",
                     k = log(n),
                     trace = 0
                     )
backwards_BIC <- step(full_model, 
                     direction = "backward", 
                     k = log(n),
                     trace = 0
                     )

model_names <- c(  "Forwards  AIC: "
                 , "Backwards AIC: "
                 , "Forwards  BIC: "
                 , "Backwards BIC: "
)

lasso_CV <- glmnet::glmnet(x = model.matrix(~ . , data = daphnia_data[train_idx, ]),
                           y = daphnia_data$LC50[train_idx],
                           family = "gaussian")

# Print formulas to compare selected models
cat(do.call(function(...) paste(..., sep = "\n")
            , as.list(diag(outer( model_names
                                  , lapply( c(  forwards_AIC$terms[[3]]
                                              , backwards_AIC$terms[[3]]
                                              , forwards_BIC$terms[[3]]
                                              , backwards_BIC$terms[[3]]
                                              )
                                            , deparse # Convert formula to string
                                            )
                                  , paste)))))
```


We see that selection based on BIC selects more sparse models. This is as expected as the BIC penalizes the dimensionality
of the model more than AIC.

We also see that backward selection chooses a more sparse models than forward selection. 

The most frugal method chose not to include H050 and C040 as predictors. The coefficients
for these variables were also not significant in the full model fit.

All the methods seem to agree on which variables are the most important.

#### Prediction error
We now compare the prediction errors of the models, computed on the test set.



```{r}

truth <- daphnia_data$LC50[test_idx]
models <- list(forwards_AIC, backwards_AIC, forwards_BIC, backwards_BIC)

for (i in 1:4) {
  predicted <- predict(models[[i]], newdata = daphnia_data[test_idx,])
  cat(paste(model_names[i], round(mean((truth - predicted)^2), 4), "\n"))
}
```
The most sparse model, chosen by backwards elimination using BIC, performed
better than the more saturated models on the test set.

Backward selection has the advantage of accounting for the effect of other 
important variables in the model when deciding on whether to keep a variable in
the model. This may be why this method chose the best model in this case.

## 4. Ridge regression

### Find tuning parameter

#### Cross validation

The number of folds in cross validation is a bias variance tradeoff.
Leave one out cross-validation has low bias but high variance, because the cross-validation
sets are so similar to each other, we are close to only making one estimation of the error.
The bias depends on sample size and the effect of sample size on model bias and variance.
Because we have relatively many observations we will choose to use 10-fold cross validation.

We now find the optimal lambda according to deviance, and plot the error estimated for
each lambda.

```{r}
library(glmnet)
set.seed(432)
x = model.matrix(LC50 ~ . - 1, data = daphnia_data[train_idx, ])
y = daphnia_data$LC50[train_idx]
cv_lambda <- cv.glmnet(x, y
                       , lambda = exp(seq(-6, 0, length.out = 100))
                       , nfolds = 10
                       , alpha = 0  # Ridge regression
                       )
plot(cv_lambda)
```

The dotted lines show the minimizing lambda, and the lambda chosen according to the "one standard
deviation rule".
The "one standard deviation rule" is to choose the most
parsimonious model parameter less than one standard deviation from the minimizing value.


```{r}
print(cv_lambda$lambda.1se)
print(mean(cv_lambda$cvsd))
```
Cross validation chooses $\lambda = 0.42$. The estimated standard deviation of the cross validation estimate
is $0.19$.


#### Bootstrapping
We will now estimate the optimal $\lambda$ with respect to deviance using bootstrapping.
```{r}

# The chosen grid of lambda-values, to optimize over
lda_seq <- exp(seq(-6, 0, length.out = 100))

deviance_estimate <- NULL
error_SD_estimate <- NULL

for (lda in 1:100) {
  MSE_boot <- NULL
  for (i in 1:100) {
    id_boot <- sample(1:nrow(daphnia_data), replace = TRUE)
    
    # Select bootstrap sample
    x = model.matrix(LC50 ~ . -1, data = daphnia_data[id_boot, ])
    y = daphnia_data$LC50[id_boot]
    # Store mean squared prediction error
    MSE_boot[i] <- mean( (predict(glmnet(x, y, family = "gaussian", alpha = 0, 
                                         lambda = lda_seq[lda])
                                  , newx = model.matrix(LC50 ~ . -1, data = daphnia_data[-id_boot, ])
                                  , newy = daphnia_data$LC50[-id_boot]
                                  )
                          - daphnia_data$LC50[-id_boot])^2
                         )
  }
  # Average over error for this lambda
  deviance_estimate[lda] <- mean(MSE_boot)
  # Estimate standard deviation of errors.
  error_SD_estimate[lda] <- sd(MSE_boot)
}
```



We plot the estimated expected prediction error for each lambda. 
```{r}
plot(log(lda_seq), deviance_estimate, xlab=expression(log(lambda)),
     ylab="Mean-squared error")
lines(log(lda_seq), deviance_estimate + error_SD_estimate, lty=2, col="red")
lines(log(lda_seq), deviance_estimate - error_SD_estimate, lty=2, col="red")
id_lowest_error <- which.min(deviance_estimate)
one_sd_from_min <- deviance_estimate[id_lowest_error] + error_SD_estimate[id_lowest_error]
abline(h=one_sd_from_min,
       lty=1, col="grey")

# Find lambda according to "one standard deviation rule"
boot_lda <- max(lda_seq[deviance_estimate < one_sd_from_min])
abline(v=log(lda_seq[id_lowest_error]), lty=3, col="blue")
abline(v=log(boot_lda), lty=3, col="blue")
```


The red bands are estimated standard deviation of the estimated error. The blue lines shows the minimizing lambda and
the lambda chosen by the "one standard deviation rule".



```{r}
print(boot_lda)
print(mean(error_SD_estimate))
```
The bootstrapping procedure chooses $\lambda = 0.32$. The estimated standard deviation of
the bootstrap estimates is $0.16$.

In order to compare the two procedures we now superimpose the two plots of estimated errors.
```{r}
plot(cv_lambda)
points(log(lda_seq), deviance_estimate, xlab=expression(log(lambda)),
       ylab="Mean-squared error")
lines(log(lda_seq), deviance_estimate + error_SD_estimate, lty=2, col="red")
lines(log(lda_seq), deviance_estimate - error_SD_estimate, lty=2, col="red")
id_lowest_error <- which.min(deviance_estimate)
one_sd_from_min <- deviance_estimate[id_lowest_error] + error_SD_estimate[id_lowest_error]
abline(h=one_sd_from_min,
       lty=2, col="grey")

# Find lambda according to "one standard deviation rule"
boot_lda <- max(lda_seq[deviance_estimate < one_sd_from_min])
abline(v=log(lda_seq[id_lowest_error]), lty=3, col="blue")
abline(v=log(boot_lda), lty=3, col="blue")
```
We see that in this case the procedures obtained a similar result. There is
still some difference, and it could have been higher. There is much
variability in each procedure, and the somewhat arbitrary
choice of number of folds and number of bootstrap iterations will effect this.


## 5. Generalized additive models
We load the required package.
```{r}
library(gam)
```

And fit a relatively complex GAM with smoothing splines for each variable. The complexity parameter, effective degrees
of freedom of the splines is set to five, a moderately complex model. Effective degrees of freedom is
linked to the penalty on the second derivative of the smoothing splines.
One degree of freedom produces an almost linear fit.
```{r}
# Fit the GAM
fit_GAM <- gam(LC50 ~ s(TPSA, 5)
           + s(SAacc, 5)
           + s(H050, 5)
           + s(MLOGP, 5) 
           + s(RDCHI, 5) 
           + s(GATS1p, 5) 
           + s(nN, 5) 
           + s(C040, 5)
           , family = gaussian
           , data = daphnia_data[train_idx, ])
summary(fit_GAM)
```
We see that, except for C040, all parameters have a significant effect either in the parametric
or in the nonparametric effect, where the nonparametric effect is simply the linear effect.

We now plot the parameters, to inspect the nonlinearity of the estimated effects.
```{r}
par(mfrow=c(2,4))
plot(fit_GAM, se=T)
```


The algorithm does not seem to fit very nonlinear effects at the bulk of the observations.
There might be a slight nonlinear effect for the categorical variables H050, nN and C040.
Perhaps this would better be modeled by factor variables, but at the cost of added complexity and
risk of some overfitting.

We now fit a less complex GAM for comparison. We set the estimated degrees of freedom for the splines to one.

```{r}
# Fit the GAM
linear_GAM <- gam(LC50 ~ s(TPSA, 1.1)
           + s(SAacc, 1.1)
           + s(H050, 1.1)
           + s(MLOGP, 1.1) 
           + s(RDCHI, 1.1) 
           + s(GATS1p, 1.1) 
           + s(nN, 1.1) 
           + s(C040, 1.1)
           , family = gaussian
           , data = daphnia_data[train_idx, ])
summary(linear_GAM)

```
There is not much increase in AIC from the more complex model, indicating that there is not much gained from the
nonlinear effects. It might appear that C040 now has a significant linear effect, but if we inspect the plot of the
parameter, we see that the parameter is estimated to be near 0 for all values of C040.

Plots
```{r}
par(mfrow=c(2,4))
plot(linear_GAM, se=T)
```


## 6. Regression trees

We will use the `rpart` package to compute regression trees. This package uses cross validation to estimate 
```{r}
library(rpart)
```


```{r}
fitted_tree <- rpart(LC50 ~ ., data = daphnia_data[train_idx, ], method = "anova",
                     control = rpart.control(xval = 10)  # 10-fold cross validation for pruning
                     )
```


```{r}
rsq.rpart(fitted_tree)
# rpart.control
# rpart.object
# .summary
# .print
```

The algorithm has chosen not to use the count variables nN, and C040. I could not find out why,
as it has chosen to keep H050, which is also a count variable.

From the plot of improvement in R-square, and reduction in relative error we see
that the biggest improvement is in the first two splits.

We now use 10-fold cross validation to choose the optimal tuning parameter `cp` according to
mean squared error.

```{r}
library(caret)
```
```{r}
set.seed(432)
fitControl <- trainControl(## 10-fold CV
                           method = "repeatedcv",
                           number = 10,
                           ## repeated ten times
                           repeats = 10)
tree_fit <- train(LC50 ~ ., data = daphnia_data[train_idx, ], 
                 method = 'rpart',
                 trControl = fitControl,
                 ## This last option is actually one
                 ## for gbm() that passes through
                 )
tree_fit
```


Cross validation chooses `cp = 0.05466436`. We prune the model using this complexity parameter.


```{r}
pruned <- rpart::prune(fitted_tree, cp = 0.05466436)
# Uses the `rpart.plot` package to plot the tree properly
rpart.plot::prp(pruned, digits = 4)
```


The selected model only uses two splits, as expected.

## 7. Comparing all the models

```{r}
predicted_train <- list(  predict(fit_linear, newdata = daphnia_data[train_idx, ])
                  , predict(backwards_BIC, newdata = daphnia_data[train_idx, ])
                  , predict(fit_GAM, newdata = daphnia_data[train_idx, ])
                  , predict(pruned, newdata = daphnia_data[train_idx, ])
                  , predict(cv_lambda,
                            newx = model.matrix(LC50 ~ . - 1, data = daphnia_data[train_idx, ]),
                            newy = daphnia_data$LC50[train_idx])
                  )
predicted_test <- list(  predict(fit_linear, newdata = daphnia_data[test_idx, ])
                       , predict(backwards_BIC, newdata = daphnia_data[test_idx, ])
                       , predict(fit_GAM, newdata = daphnia_data[test_idx, ])
                       , predict(pruned, newdata = daphnia_data[test_idx, ])
                       , predict(cv_lambda,
                                 newx = model.matrix(LC50 ~ . - 1, data = daphnia_data[test_idx, ]),
                                 newy = daphnia_data$LC50[test_idx])
)
RSS_train <- NULL
RSS_test <- NULL
for (i in 1:5) {
  RSS_test[i] <-  mean( ( predicted_test[[i]]  - daphnia_data$LC50[test_idx]  )^2 )
  RSS_train[i] <- mean( ( predicted_train[[i]] - daphnia_data$LC50[train_idx] )^2 )
}
```
```{r}
# plot(meanResidents, axes=FALSE, xlab="dorms")
models <- c(  "Linear reg."
            , "Var. sel."
            , "GAM"
            , "tree"
            , "Ridge r."
            )
barplot(rbind(RSS_train, RSS_test), beside=T, names.arg = models,
        legend.text = c("Train", "Test"),
        )
```
We see that all the models except the regression tree perform very similarly on the test set.
The generalized linear model performs better on the training data, but performs worse
on the test set, indicating that it has overfit the data. The regression tree performs
the worst on both sets, indicating that a linear model adopted by the other models
will model the underlying mechanism well.



# Problem 2. Classification

Unfortunately I ran out of time to do this problem properly. I will go over it more thoroughly later.


```{r}
# Clear the R environment
rm(list=ls())
library(mlbench)
data(PimaIndiansDiabetes)
```

## Split data
We split the data set into a trianing and test set. The method `createDataPartition` preserves
the overall class distribution of the predictor.
```{r}
library(caret)
trainIndex <- createDataPartition(PimaIndiansDiabetes$diabetes, p = 2/3, 
                                  list = FALSE, 
                                  times = 1)

# Check if class proportions are equal
a <- summary(PimaIndiansDiabetes$diabetes)
a[1]/a[2]
a <- summary(PimaIndiansDiabetes$diabetes[trainIndex])
a[1]/a[2]
a <- summary(PimaIndiansDiabetes$diabetes[-trainIndex])
a[1]/a[2]
```

## 1. k-neares neighbours

We find the number of neighbours using 5-fold cross validation, using the caret package
for convenience.


```{r}
set.seed(432)
fitControl <- trainControl(method = "repeatedcv",
                           number = 5, # 5 folds
                            repeats = 10 # repeated ten times
                           )
knn_fit_5f <- train(diabetes ~ ., data = PimaIndiansDiabetes[trainIndex, ], 
                 method = 'knn',
                 trControl = fitControl,
                 preProcess = c("center", "scale"),
                 tuneLength = 20
                 )
knn_fit_5f
```

```{r}
set.seed(432)
fitControl <- trainControl(method = "LOOCV")
knn_fit_loo <- train(diabetes ~ ., data = PimaIndiansDiabetes[trainIndex, ], 
                 method = 'knn',
                 trControl = fitControl,
                 preProcess = c("center", "scale"),
                 tuneLength = 20
                 )
knn_fit_loo
```
```{r}

# Noramlize test and train data based on the training data, with the help of the caret package
normalized_data <- predict(preProcess(PimaIndiansDiabetes[trainIndex, ],
                                      method = c("center", "scale")),
                           newdata = PimaIndiansDiabetes)

predicted_knn <- function(k) {
  # Calculates prediction on the test set of a kNN-model
  # trained on the test set using k neighbours
  class::knn(train = normalized_data[trainIndex, -9],
             test = normalized_data[-trainIndex, -9], 
             cl = normalized_data$diabetes[trainIndex],
             k = k
             )
}

k_vals <- knn_fit_5f$results[["k"]]

test_error <- NULL
for (i in 1:length(k_vals)) {
  test_error[i] <- 1 - mean(predicted_knn(k=k_vals[i]) 
                            == PimaIndiansDiabetes$diabetes[-trainIndex])
}
```

We inspect the test error, using proportion of mispredictions as a measure.


```{r}
plot(k_vals, 1 - knn_fit_5f$results[["Accuracy"]], ylim = c(0.2, 0.3),
     col = "blue", type = "l",
     xlab = "k", ylab = "Proportion of mispredictions")
lines(k_vals, 1 - knn_fit_loo$results[["Accuracy"]], col = "red")
lines(k_vals, test_error, col = "black")
legend("bottomleft", legend = c("Test error", "Train error, 5 folds", "Train error, LOO"),
       lty = 1,
       col = c("black", "blue", "red"))
```


We see that leave-one-out CV has higher variance than 5-fold CV, as expected. Both seem to aim for the
same error estimate, but bots seem to underestimate the true error in the data.

The test error seem to be minimized for k = 12, a value far below the chosen k values, that would
fit the model more closely to the data. But this minimizing value seem to be the product of random
variation, and I would expect the "one standard deviation" rule to imply a far higher value for k,
seeng as the error does not change much for k up to 40.


## 2. GAM for classification

```{r, results=F, message=F, warning=F}

set.seed(432)
fitControl <- trainControl(method = "repeatedcv",
                           number = 10,
                           repeats = 2,
)
gam_fit <- train(diabetes ~ ., data = PimaIndiansDiabetes[trainIndex, ], 
                 method = 'gamboost',
                 # trControl = fitControl,
                 preProcess = c("center", "scale")
                 # , tuneLength = 10
)
gam_fit
```
I was not able to figure this method out properly.



## 3. Classification trees


### Bagging tree
```{r}
library(ipred)
set.seed(432)
fit_tree_bagging <- bagging(diabetes ~ ., data = normalized_data[trainIndex, ], coob = T)

fitControl <- trainControl(method = "repeatedcv",
                           number = 10,
                           repeats = 3)
tree_fit <- train(diabetes ~ ., data = PimaIndiansDiabetes[trainIndex, ], 
                 method = 'rpart',
                 trControl = fitControl,
                 )
```
```{r}
pruned_bagging <- prune(tree = fit_tree_bagging, cp = tree_fit$bestTune[[1]])
print(pruned_bagging)
```

### Random forest

```{r}
set.seed(432)
random_forest <- train(diabetes ~ ., data = PimaIndiansDiabetes[trainIndex, ], 
                 method = 'cforest',
                 trControl = fitControl,
                 preProcess = c("center", "scale")
                 # , tuneLength = 10
)
random_forest
```

### Neural network
```{r, results=F}
set.seed(432)
nnet <- train(diabetes ~ ., data = PimaIndiansDiabetes[trainIndex, ], 
                 method = 'nnet',
                 trControl = fitControl,
                 preProcess = c("center", "scale")
                 # , tuneLength = 10
)
```


```{r}
nnet
```




### Boosted tree
```{r}
set.seed(432)
ada_fit <- train(diabetes ~ ., data = PimaIndiansDiabetes[trainIndex, ], 
                 method = 'ada',
                 trControl = fitControl,
                 preProcess = c("center", "scale")
                 # , tuneLength = 10
)
ada_fit
```

## 4. Choice of method
```{r}
predicted_test <- lapply(
  list(knn_fit_5f,
       knn_fit_loo,
       gam_fit, 
       random_forest,
       nnet,
       ada_fit
  ),
  function (x) predict(x,
                       newdata = PimaIndiansDiabetes[-trainIndex, ])
)
predicted_test[[7]] <- predict( pruned_bagging, newdata = normalized_data[-trainIndex, ])

errors <- sapply(predicted_test,
                 function(x) 1- mean(x == PimaIndiansDiabetes$diabetes[-trainIndex]))

models <- c(  "kNN 5f"
            , "kNN LOO"
            , "GAM"
            , "Rnd for"
            , "nnet"
            , "ADA"
            , "Bagging"
            )
barplot(errors[-3], beside=F, names.arg = models[-3])
```


I was not able to properly implement the GAM procedure, so we will simply compare the other models.
The difference in their performance is so small that the choice is almost arbitrary. Still
the neural network performed slightly better than the rest on the test set, and is therefore
the recommended model.





