---
title: "Comparing the predictive power of Lasso and Ridge Regression"
author: "Jonas Øren"
date: "9/16/2020"
output: pdf_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries, include=FALSE}
library(naniar)
library(dplyr)
```

# Introduction
Lasso and Ridge regression are two methods for handling large dimensional data in regression analysis.
Both methods penalize large values of predictors through a penalty term in the likelihood
maximizing procedure.
Ridge regression uses a quadratic penalty
$$
  \hat \beta^{\text{ridge}}=\text{argmin}\Big\{\sum_{i=1}^n\big( Y_i-\beta_0 +
  \sum_{j=1}^pX_{ij}\beta_j \big)^2 + \lambda 
  \sum_{j=1}^p \beta^2_p
  \Big\}.
$$
Lasso on the other hand penalizes the absolute value of the parameters
$$
  \hat \beta^{\text{lasso}}=\text{argmin}\Big\{\sum_{i=1}^n\big( Y_i-\beta_0 +
  \sum_{j=1}^pX_{ij}\beta_j \big)^2 + \lambda 
  \sum_{j=1}^p |\beta_p|
  \Big\}
$$
The penalty boundary of the absolute value function will usually cause some parameter estimates to be zero,
leading to automatic variable selection as well.

The goal of this analysis is to test the predictive power of Lasso vs Ridge Regression on the data of crime rates
in selected American cities.


# About the data

The data is collected to describe the people in each area, their living
situation and economic situation. For example `racePctHisp`: percentage of
population that is of hispanic, `perCapInc`: per capita income, and `pctUrban`:
percentage of people living in areas classified as urban.

The covariates and response variable come normalized between zero and one.
Certain extreme outliers are truncated to zero or one manually. For details see
the `communities.names` file.


## Missing data
Many of the variables have much missing data. We simply remove variables with a lot of
missing data, as they are not useful to us.


## Collinearity
We will not check variables for colinearity because we will be using penalizing
methods that mitigate the effect of colinearity by preventing large opposite
estimated values for correlated variables. In addition, as our focus is
prediction and not interpretation colinearity is not as big a concern in this
case.

# Model

We assume the outcomes are a linear function of the covariates, plus noise,

$$
Y_i = \beta_0 + \sum_{j=1}^p \beta_j X_j + \epsilon
$$
We will check if this assumption holds by inspecting the residuals, and then
assess the models prediction accuracy by splitting the data into a training- and
test set, and check prediction accuracy on the data not used for model fitting.
This is to ensure the prediction evaluation is not biased by the correlation
between the training data
and the fitted parameters.

# Analysis
We read the data and give the parameters their correct names.
```{r data}
data = read.csv("./data/communities.data")
# Set correct names for variables
names(data) = c("state", "county", "community", "communityname", "fold", "population",
"householdsize", "racepctblack", "racePctWhite", "racePctAsian", "racePctHisp",
"agePct12t21", "agePct12t29", "agePct16t24", "agePct65up", "numbUrban", "pctUrban",
"medIncome", "pctWWage", "pctWFarmSelf", "pctWInvInc", "pctWSocSec", "pctWPubAsst",
"pctWRetire", "medFamInc", "perCapInc", "whitePerCap", "blackPerCap", "indianPerCap",
"AsianPerCap", "OtherPerCap", "HispPerCap", "NumUnderPov", "PctPopUnderPov",
"PctLess9thGrade", "PctNotHSGrad", "PctBSorMore", "PctUnemployed", "PctEmploy",
"PctEmplManu", "PctEmplProfServ", "PctOccupManu", "PctOccupMgmtProf", "MalePctDivorce",
"MalePctNevMarr", "FemalePctDiv", "TotalPctDiv", "PersPerFam", "PctFam2Par",
"PctKids2Par", "PctYoungKids2Par", "PctTeen2Par", "PctWorkMomYoungKids", "PctWorkMom",
"NumIlleg", "PctIlleg", "NumImmig", "PctImmigRecent", "PctImmigRec5", "PctImmigRec8",
"PctImmigRec10", "PctRecentImmig", "PctRecImmig5", "PctRecImmig8", "PctRecImmig10",
"PctSpeakEnglOnly", "PctNotSpeakEnglWell", "PctLargHouseFam", "PctLargHouseOccup",
"PersPerOccupHous", "PersPerOwnOccHous", "PersPerRentOccHous", "PctPersOwnOccup",
"PctPersDenseHous", "PctHousLess3BR", "MedNumBR", "HousVacant", "PctHousOccup",
"PctHousOwnOcc", "PctVacantBoarded", "PctVacMore6Mos", "MedYrHousBuilt", "PctHousNoPhone",
"PctWOFullPlumb", "OwnOccLowQuart", "OwnOccMedVal", "OwnOccHiQuart", "RentLowQ",
"RentMedian", "RentHighQ", "MedRent", "MedRentPctHousInc", "MedOwnCostPctInc",
"MedOwnCostPctIncNoMtg", "NumInShelters", "NumStreet", "PctForeignBorn",
"PctBornSameState", "PctSameHouse85", "PctSameCity85", "PctSameState85", "LemasSwornFT",
"LemasSwFTPerPop", "LemasSwFTFieldOps", "LemasSwFTFieldPerPop", "LemasTotalReq",
"LemasTotReqPerPop", "PolicReqPerOffic", "PolicPerPop", "RacialMatchCommPol",
"PctPolicWhite", "PctPolicBlack", "PctPolicHisp", "PctPolicAsian", "PctPolicMinor",
"OfficAssgnDrugUnits", "NumKindsDrugsSeiz", "PolicAveOTWorked", "LandArea", "PopDens",
"PctUsePubTrans", "PolicCars", "PolicOperBudg", "LemasPctPolicOnPatr",
"LemasGangUnitDeploy", "LemasPctOfficDrugUn", "PolicBudgPerPop", "ViolentCrimesPerPop")
```


## Missing data
We then remove variables with a lot of missing data, meaning variables that have
more than 100 missing observations. There are still some observations with
missing data. We will remove these. As there is only one observation left with
missing data, wchich is 0.05% of the total number of observations, we will
remove it without fear of biasing the predictions.


```{r clean_data}
# Change missing values to NA data type
data <-  naniar::replace_with_na_all(data, condition = ~.x == "?" )

# Drop columns with more than 100 NA values
omitted <- (names(data[, colSums(is.na(data)) > 100]))
data <- data[, colSums(is.na(data)) < 100]
cat(paste("Omitting variables with > 100 missing data. Omitted:\n",
          paste(omitted, collapse = "\n ")))

# Remove observations with missing data
n_omitted <- nrow(data)-nrow(na.omit(data))
omitted <- paste(names(data[, colSums(is.na(data)) > 0]), collapse = "\n ")
data <- na.omit(data)
cat(paste("Omitting observations with missing data. Number of obs. removed: ",
          n_omitted,
          "\n-Variables that contained missing data: \n",
          omitted
          ))
```


We then ensure that all variables have the correct datatype. All variables are numerical
```{r}
# Convert to numeric datatype (covariates only)
data[-c(1,2,3)] <- lapply(data[-c(1,2,3)], function(x) {
    if(is.factor(x)) as.numeric(as.character(x)) else x
})
```

Finally, we store the response variable and covariates for later use
```{r}
## Store variables
response <- data$ViolentCrimesPerPop
# Remove first five variables, and the response variable
covariates <- data[-c(1,2,3, which(names(data)=="ViolentCrimesPerPop"))]
```

## Split into train and test data
We then spilt the data into a training- and test test, in order to test the predictive power of our models
on new data not used in the fitting process.

```{r test_train}
set.seed(432)   # For reproducability

# Indexes
n <- nrow(data)
train <- sample(1:n, size = round(n*0.75), replace = FALSE)
test <- (1:n)[-train]
```


## Fitting the models
We set up a general function for finding the tuning parameter and fitting a model.
The tuning parameter $\lambda$ will be chosen using cross-validation, by the `cv.glmnet`-function. The set was
be chosen by trial and error to ensure it contains a minimizing point.
The model will be fitted by `glmnet` where the parameter `alpha` decides if it is a lasso or ridge-regression
fitting procedure.

```{r lasso}
glmnet_analysis <- function(alpha) {
  ## Perform glmnet analysis
  ## alpha = 1  -> lasso
  ## alpha = 0  -> ridge regression

  mm <- as.matrix(covariates[train, ])  
  
  # Find lambda with leave one out cross validation
  cv_fit <- glmnet::cv.glmnet(x = mm,
                              y = response[train],
                              alpha = alpha,
                              lambda = exp(seq((-10), (-2), length.out = 100))
  )
  
  # Fit model using the best lambda
  fit <- glmnet::glmnet(x = mm,
                        y = response[train],
                        alpha = alpha,
                        lambda = cv_fit$lambda.min
  )
  # predict(fit, newx=as.matrix(covariates[test, ]))
  predicted <- predict(fit, newx=as.matrix(covariates[test, ]))[,1]
  error <- predicted-response[test]
  return(list(error=error, cv_fit=cv_fit, fit=fit))
}
```

## Fit the models
We then fit the models
```{r}
lasso <- glmnet_analysis(alpha = 1)
ridge <- glmnet_analysis(alpha = 0)
```

## Lambda
Now we must inspect the cross-validation mean squared error for the sequence of lambdas, to make sure
a good value was chosen.
```{r}
plot(lasso$cv_fit)
plot(ridge$cv_fit)
```
The plots indicate that the procedure has found reasonable $\lambda$ values. The chosen log($\lambda$) values were

```{r}
cat(paste(
  "Lasso log-lambda: ", 
  round(log(lasso$fit$lambda), 2),
  "\nRidge Regression log-lambda: ", 
  round(log(ridge$fit$lambda),2)
))
```

## Check linearity assumption
We then inspect the residuals, to see if they look linear.

```{r}
plot(response[test], lasso$error)
plot(response[test], ridge$error)
```

The residuals do indeed appear to have a linear relationship with the response. This supports our assumption
of a linear model.


## Coefficients
We now inspect the coefficients chosen by the two methods, by printing the 20 largest chosen coefficients in absolute value.
```{r}
ridge_coef <- coef(ridge$fit)
lasso_coef <- coef(lasso$fit)
ridge_varname <- head(ridge_coef@Dimnames[[1]][order(abs(ridge_coef@x), decreasing = T)], 20)
ridge_var <- head(sort(abs(ridge_coef@x), decreasing = T), 20)

index_of_lasso_coefs <- head((lasso_coef@i+1)[order(lasso_coef@x, decreasing = T)], 20)
lasso_varname <- head(lasso_coef@Dimnames[[1]][index_of_lasso_coefs], 20)
lasso_var <- head(sort(lasso_coef@x, decreasing = T), 20)
data.frame(
  lasso_varname = lasso_varname,
  lasso_var = lasso_var,
  ridge_varname = ridge_varname,
  ridge_var = ridge_var
)
```

We see that the two methods choose roughly the same variables to be the most important.
The lasso method has discarded 36 of the 101 variables.
```{r}
lasso_coef@Dim[1] - lasso_coef@p[2]
```


## Prediction efficiency
We will assess the prediction efficiency by two metrics. First the mean absolute error (MAE)
$$
\frac{1}{n}\sum_{i=1}^n|\hat Y_i - Y_i|
$$
where $\hat Y_i$ is the prediction.
Because the response variable ''violent chrime per capita'' is normalized between zero and one we interpret
absolute error as how close the prediction is in terms of the range of "reasonable" values. If the error is larger than one, we are far outside of reasonable predictions. The first and third quantiles are 0.07 and 0.33.
```{r}
summary(response)
```

This  is the range of the bulk of the response variable. As we see from the histogram of the response.

```{r}
hist(response)
```

Second we will inspect the mean squared error (MSE)
$$
\frac{1}{n}\sum_{i=1}^n(\hat Y_i - Y_i)^2
$$
The MSE is heavily influenced by large errors. If there is a large difference between MAE and MSE we deduce
that there there are some errors that are very large compared to the MAE.

```{r}
n <- length(test)
cat(paste(
  "MAE lasso:            ",
  round(sum(abs(lasso$error))/n, 5),
  "\nMAE Ridge Regression: ",
  round(sum(abs(ridge$error))/n, 5),
  "\nMSE lasso:            ",
  round(sqrt(sum((ridge$error^2))/n), 5),
  "\nMSE Ridge Regression: ",
  round(sqrt(sum((lasso$error^2))/n), 5)
))
```

On average both methods predictions are about 10% off the mark. But Ridge regression seems to perform slightly better.
From MSE we see that there does not seem to be many extreme errors, but here lasso seem to make somewhat less relatively large errors.

# Conclution
Ridge regression and Lasso both handle seem to handle prediction of crime rates reasonably well. For prediction the choice between the two
is nearly arbitrary, as they performed very similarly. If one is to be recommended the author would suggest lasso, as it seems to be slightly more robust
in that it makes less large errors.
